Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-17 ieee_paper.tex
Added 2 citations and mods to figures.

\section{Introduction}
The multiplication of sparse matrix by a dense vector (SpMV) is an
important kernel in many fields from scientific computing~\cite{} \todo{Do you have a good citation for that?}
to graph analysis (recommender systems~\cite{Brin98}, graph
drawing~\cite{Koren05}). So improving the performance of SpMV has
captured the interest of many researcher; including the development of
various implementations for multi-core
CPUs~\cite{Buluc2009SPAA,Williams07} and GPUs~\cite{Bell08,
  conf/ipps/KreutzerHWFBB12,
  journals/concurrency/VazquezFG11,kumar2012accelerating}. The main
challenge that is faced to obtain good performance for SpMV is that
the operation are conducted using memory location that are irregular
and often unpredictible. That makes the kernels mostly memory
bound and there is a significant instruction overhead per flop.

@@

An other solution, and the one we pursue in this paper, is to compute
multiple SpMV at once on matrices that have the same sparsity
patterns. Obviously not all the applications have such a property. But
important classes of applications such as graph
recommendation~\cite{Kucuktunc13-SNAM}, eigensolving~\cite{LOBPCG} and the computation of
derivative of Radial Basis Functions(RBF)~\cite{} \todo{a good citation?} can use multiple
SpMVs simultaneously. In this paper in particular, we investigate the
case of the derivative of RBFs where four derivatives of four
different functions is expressed as the multiplication of four vectors
by four matrices with identical sparsity patterns leading to the
simultaneous execution of 16 SpMVs at a time. Using multiple vectors
at a time has been investigated before to densify the
computations~\cite{Im01} but this is the first time that the
densification comes from adding both vectors and matrices.

@@

  \caption{Estimation of the reachable performance one can achieve
    using a device with a memory to computational units of 150GB/s
    when varying the number of vectors and matrices. \todo{we could
      remove the speedup plot if we need space.}  \todo{can the legend
      read "Single Precision (best)"} \todo{there is a missing
      parenthesis in the title; alternatively remove title.}}
  \label{fig:perf_predict}
\end{figure*}

@@

%<<<<<<< HEAD
%\begin{figure}
%  \centering
%  \includegraphics[width=.45\textwidth]{figures/plot_for_natasha_ncar_end_of_year.pdf}
%
%  \caption{Performance on a MIC of derivative computation for RBFFD
%    with 884,736 nodes distributed quasi-randomly in a cube. Single
%    derivative of a single function (base case, triangle) and four
%    derivatives of four functions (multi case, circles). No bandwidth
%    reduction (blue) and Reverse Cuthill-McKee (red).\todo{should we have a similar one for the host?}}
%\end{figure}
%
\begin{figure*}
\centering
\subfigure[Best case.]{\includegraphics[width=.33\linewidth]{figures/flops_to_bytes_best-crop.pdf}} %\label{fig:ratio_best}}
\subfigure[Worst case.]{\includegraphics[width=.33\linewidth]{figures/flops_to_bytes_worst-crop.pdf}}%\label{fig:ratio_worst}}
\subfigure[Worst case (No cacheline effects)]{\includegraphics[width=.33\linewidth]{figures/flops_to_bytes_no_cache-crop.pdf}}%\label{fig:ratio_nocache}}
\caption{Ratio of flops to bytes; (a) simple count of flops and bytes that require transfer from memory;(b) best and (c) worst
case taking into account that bringing one element of $x$ into memory brings along the remainder of the cache line.}
\label{fig:ratio_bytes_flops}
\end{figure*}
%=======
%\begin{figure}
%  \centering
%  \includegraphics[width=.45\textwidth]{figures/plot_for_natasha_ncar_end_of_year.pdf}
%
%  \caption{Performance on a MIC of derivative computation for RBFFD
%    with 884,736 nodes distributed quasi-randomly in a cube. Single
%    derivative of a single function (base case, triangle) and four
%    derivatives of four functions (multi case, circles). No bandwidth
%    reduction (blue) and Reverse Cuthill-McKee (red).\todo{should we have a similar one for the host?}}
%\end{figure}
%
%>>>>>>> 619115cea6da9466f09a3b01456eacf7991606e2

>>>>

\section{Introduction}
The multiplication of sparse matrix by a dense vector (SpMV) is an
important kernel in many applied fields such as fluid dynamics 
\cite{journals/tog/BolzFGS03} 
to graph analysis (recommender systems~\cite{Brin98}, graph
drawing~\cite{Koren05}). So improving the performance of SpMV has
captured the interest of many researcher; including the development of
various implementations for multi-core
CPUs~\cite{Buluc2009SPAA,Williams07} and GPUs~\cite{Bell08,
  conf/ipps/KreutzerHWFBB12,
  journals/concurrency/VazquezFG11,kumar2012accelerating}. The main
challenge that is faced to obtain good performance for SpMV is that
the operation are conducted using memory location that are irregular
and often unpredictible. That makes the kernels mostly memory
bound and there is a significant instruction overhead per flop.

@@

An other solution, and the one we pursue in this paper, is to compute
multiple SpMV at once on matrices that have the same sparsity
patterns. Obviously not all the applications have such a property. But
important classes of applications such as graph
recommendation~\cite{Kucuktunc13-SNAM}, eigensolving~\cite{LOBPCG} and the computation of
derivatives of Radial Basis Functions(RBF)~\cite{FLBWSC12} can use multiple
SpMVs simultaneously. In this paper in particular, we investigate the
case of the derivative of RBFs where four derivatives of four
different functions is expressed as the multiplication of four vectors
by four matrices with identical sparsity patterns leading to the
simultaneous execution of 16 SpMVs at a time. Using multiple vectors
at a time has been investigated before to densify the
computations~\cite{Im01} but this is the first time that the
densification comes from adding both vectors and matrices.

@@

  \caption{Estimation of the reachable performance one can achieve
    using a device with a memory to computational units of 150GB/s
    when varying the number of vectors and matrices. \todo{we could
      remove the speedup plot if we need space.}}
  \label{fig:perf_predict}
\end{figure*}

@@

%<<<<<<< HEAD
%\begin{figure}
%  \centering
%  \includegraphics[width=.45\textwidth]{figures/plot_for_natasha_ncar_end_of_year.pdf}
%
%  \caption{Performance on a MIC of derivative computation for RBFFD
%    with 884,736 nodes distributed quasi-randomly in a cube. Single
%    derivative of a single function (base case, triangle) and four
%    derivatives of four functions (multi case, circles). No bandwidth
%    reduction (blue) and Reverse Cuthill-McKee (red).\todo{should we have a similar one for the host?}}
%\end{figure}
%
%\begin{figure*}
%\centering
%\subfigure[Best case.]{\includegraphics[width=.33\linewidth]{figures/flops_to_bytes_best-crop.pdf}} %\label{fig:ratio_best}}
%\subfigure[Worst case.]{\includegraphics[width=.33\linewidth]{figures/flops_to_bytes_worst-crop.pdf}}%\label{fig:ratio_worst}}
%\subfigure[Worst case (No cacheline effects)]{\includegraphics[width=.33\linewidth]{figures/flops_to_bytes_no_cache-crop.pdf}}%\label{fig:ratio_nocache}}
%\caption{Ratio of flops to bytes; (a) simple count of flops and bytes that require transfer from memory;(b) best and (c) worst
%case taking into account that bringing one element of $x$ into memory brings along the remainder of the cache line.}
%\label{fig:ratio_bytes_flops}
%\end{figure*}
%=======
%\begin{figure}
%  \centering
%  \includegraphics[width=.45\textwidth]{figures/plot_for_natasha_ncar_end_of_year.pdf}
%
%  \caption{Performance on a MIC of derivative computation for RBFFD
%    with 884,736 nodes distributed quasi-randomly in a cube. Single
%    derivative of a single function (base case, triangle) and four
%    derivatives of four functions (multi case, circles). No bandwidth
%    reduction (blue) and Reverse Cuthill-McKee (red).\todo{should we have a similar one for the host?}}
%\end{figure}
%
%>>>>>>> 619115cea6da9466f09a3b01456eacf7991606e2
